---
title: "How objects are found by package functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How objects are found by package functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Consider a package **teeny** that exports function
`mean_and_variance()` for calculating the sample mean and the sample
variance in a single call:

```{r}
#' Calculate the Mean and Variance
#'
#' @param x A numeric vector.
#'
#' @return
#' A named numeric vector of length two
#' with elements `mean` and `variance`.
#'
#' @importFrom stats var
#' @export
mean_and_variance <- function(x) {
  mu <- mean(x)
  sigma2 <- var(x)
  c(mean = mu, variance = sigma2)
}
```

There are a few things to notice about the implements.  The first is
that we declare that `variance()` should be imported from the
**stats** package. Here we use a **[roxygen2]** `@importFrom stats
var` statement to declare this import, which is short for manually
adding an `importFrom(stats, var)` entry in the package `NAMESPACE`
file.

The second is that, because all objects exported by the **base**
package are always available, we do not have to explicitly import
them.  It is actually not possible to import **base** functions; if we
would add an `@importFrom base mean` statement, the package fails to
_install_ (see Appendix).

Continuing, when R evaluates

```r
teeny::mean_and_variance(1:100)
```

```
##     mean variance 
##  50.5000 841.6667
```

it will eventually have to locate functions `mean()` and `var()`.  It
turns out that these are effectively found as:

```r
pkg_ns <- getNamespace("teeny")
mean <- parent.env(parent.env(pkg_ns))[["mean"]]
var  <- parent.env(           pkg_ns )[["var"]]
```

Note the extra layer of `parent.env()` for `mean()`.  So, why are they
not found in the same environment?  This is because we explicitly
imported `stats::var()`, but not `base::mean()`.

Let's dig in deeper to see how this works, but before doing that,
let's look at what the first few parent environments of the **teeny**
namespace are. We can du this using the **environments** package as:

```r
pkg_ns <- getNamespace("teeny")
nss <- environments::parent_envs(pkg_ns, until = globalenv())
nss
```

```
## $teeny
## <environment: namespace:teeny>
## 
## $`imports:teeny`
## <environment: 0x55cb372dd980>
## attr(,"name")
## [1] "imports:teeny"
## 
## $base
## <environment: namespace:base>
## 
## $R_GlobalEnv
## <environment: R_GlobalEnv>
```

This reveals (part of) the search path used by the functions in the
**teeny** package.  When `mean_and_variance()` is called, R searches
for `mean()` and `var()` in these environments until found.  So, in
the case of `mean()`, it first looks `nss[[1]]` using something like
`exists("mean", mode = "function", envir = nss[[1]], inherits =
FALSE)`, and if not found, it continues to environment `nss[[2]]`, and
so on until found.  In this case, it locates `mean()` in `nss[[3]]`,
i.e. in the **base** namespace;

```r
exists("mean", mode = "function", envir = nss[[3]], inherits = FALSE)
```

```
## [1] TRUE
```

It searches for `var()` in a similar manner.  Since `var()` was
explicitly imported, and all package imports are stored in the
`imports::teeny` environment, it finds `var()` in `nss[[2]]`;

```r
exists("var", mode = "function", envir = nss[[2]], inherits = FALSE)
```

```
## [1] TRUE
```

This begs the question, isn't that imported `var()` function a copy of
the `var()` in the **stats** package?  Yes, it is!  Technically, we
could, with some hacks, replace the "imported" `var()` with another
function after the package has been loaded.


## Forgetting to import a function

Now, what would happen if we forget to import `stats::var()`?  Let's
retry by dropping:

```r
#' @importFrom stats var
```

First of all, `R CMD check` would detect this and produce the
following NOTE:

```
* checking R code for possible problems ... NOTE
mean_and_variance: no visible global function definition for ‘var’
Undefined global functions or variables:
  var
Consider adding
  importFrom("stats", "var")
to your NAMESPACE file.
```

That's comforting to know. Note also how it suggests how we might be
able to resolve it. That is a useful service.

Next, let's see what happens if we try to use the function;

```r
teeny::mean_and_variance(1:100)
```

```
##     mean variance 
##  50.5000 841.6667
```

Hmm, how is that possible?  The `var()` function does not exist in the
package namespace, not in the "imports" environments, not in the
**base** namespace, and not in the global environment.  What we didn't
say above is that, if a function cannot be found even in the global
environment, it continues to search the parent environments of the
global environment too.  The parent environments of the global
environment are all the environments that `search()` reports;

```{r}
search()
```

If the object search for is not found in one of these environments,
then it ends up at the very top parent environment, which is always
the empty environment, and stops there (with an error).

To see _all_ the environment searched, we can use:

```r
pkg_ns <- getNamespace("teeny")
nss <- environments::parent_envs(pkg_ns, until = emptyenv())
str(nss)
```

```
## List of 13
##  $ teeny            :<environment: namespace:teeny> 
##  $ imports:teeny    :<environment: 0x555840d33fc8> 
##   ..- attr(*, "name")= chr "imports:teeny"
##  $ base             :<environment: namespace:base> 
##  $ R_GlobalEnv      :<environment: R_GlobalEnv> 
##  $ package:stats    :<environment: package:stats> 
##   ..- attr(*, "name")= chr "package:stats"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/stats"
##  $ package:graphics :<environment: package:graphics> 
##   ..- attr(*, "name")= chr "package:graphics"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/graphics"
##  $ package:grDevices:<environment: package:grDevices> 
##   ..- attr(*, "name")= chr "package:grDevices"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/grDevices"
##  $ package:utils    :<environment: package:utils> 
##   ..- attr(*, "name")= chr "package:utils"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/utils"
##  $ package:datasets :<environment: package:datasets> 
##   ..- attr(*, "name")= chr "package:datasets"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/datasets"
##  $ package:methods  :<environment: package:methods> 
##   ..- attr(*, "name")= chr "package:methods"
##   ..- attr(*, "path")= chr "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/methods"
##  $ Autoloads        :<environment: 0x55583eee63c0> 
##   ..- attr(*, "name")= chr "Autoloads"
##  $ package:base     :<environment: base> 
##  $ R_EmptyEnv       :<environment: R_EmptyEnv>
```

Thus, the reason why `teeny::mean_and_variance(1:100)` works, is that
R eventually locates `var()` in the **stats** package, i.e. in the
`package:stats` environment.  Using the **environments** package, we
can see where it finds `var()` by calling:

```r
environments::find_object(name = "var", from = teeny::mean_and_variance)
```

```
## $name
## [1] "var"
## 
## $envir
## <environment: package:stats>
## attr(,"name")
## [1] "package:stats"
## attr(,"path")
## [1] "/home/hb/shared/software/CBI/R-4.2.1-gcc9/lib/R/library/stats"
```

So, why do we even bother to import `var()` from the **stats** package
then?  Recall how also the global environment is part of the search
path.  This means that we can inject another, possible false,
implementation of `var()` by adding it to the global environment, e.g.

```r
var <- function(x) pi
teeny::mean_and_variance(1:100)
```

```
##      mean  variance 
## 50.500000  3.141593
```

Whoops!  That's really worrying; the result of the function depends on
what happens to be in the global environment of the current R session.
Furthermore, the **stats** package may not even be attached, so we
could end up with an error also a fresh R session, e.g.

```r
$ R_DEFAULT_PACKAGES=base R --quiet --vanilla
> search()
[1] ".GlobalEnv"   "Autoloads"    "package:base"
> teeny::mean_and_variance(1:100)
Error in var(x) : could not find function "var"
```

This illustrates why it is important to declare _all_ imports, even
those from base-R packages. So, do _not_ ignore those NOTEs on "no
visible global function definition for ..." that `R CMD check` reports
on.



# Appendix

## Why R searches also the global environment and attached packages

The parent environment of the **base** namespace is the global environment;

```{r}
base_ns <- getNamespace("base")
environments::parent_envs(base_ns)
```

So, why isn't it just the empty environment?  Then we would avoid the
problem of picking up unwanted objects in the global environment, and
avoid being dependent on what packages happens to be attached at the
time we call our package functions.  
The answer has to do with historical reasons.  The design of having
the **base** namespace being followed by the global environment
stems from the old days when R did not have namespaces.  At that time,
the R engine searched all attached packages to identify any variable
or function used to evaluate a function.  That design had of course
similar problems to the ones illustrated above, e.g. functions could
be overridden by assign new ones in the global environment, or by a
third-party package attached on the `search()` path.  This is why the
concept of _namespaces_ was introduced in R; as a package developer
you had full control of exactly which implementation of a function was
called.

However, due to how S3 method dispatching worked, and how some
packages attaches packages conditionally of being available, e.g. `if
require(otherpkg)) somefcn()`, it was necessary to keep searching also
the global environment and attached packages as a fallback.  These are
the main reasons why the **base** namespace is still followed by the
global environment.  It is not unreasonable to expect that this will
change in some future R version.  For instance, in the R-devel thread
'Time to drop globalenv() from searches in package code?' (2022-09-15;
<https://stat.ethz.ch/pipermail/r-devel/2022-September/081980.html>),
one of the R Core developers explains the history and proposes a way
forward to make the empty environment the parent of the **base**
namespace.


## The difference between getNamespace("base") and baseenv()

If we look at the output from, for instance, `parent_envs(stats::median)`, we see that there are two "base" environments.  One early on following the "imports" environment and one at the very end just before the empty environment.  These two environments are:

```{r}
getNamespace("base")
```

and

```{r}
baseenv()
```

respectively.  They are very similar, except that their parent environments differ; `getNamespace("base")` is followed by the _global_ environment, and `baseenv()` is followed by the _empty_ environment;

```{r}
environments::parent_env(getNamespace("base"))

environments::parent_env(baseenv())
```

Everything else is the same, e.g.

```{r}
all.equal(getNamespace("base"), baseenv())
```

and

```{r}
identical(getNamespace("base")[["mean"]], baseenv()[["mean"]])
```

By the way, `.BaseNamespaceEnv` is the same as `getNamespace("base")`;

```{r}
identical(.BaseNamespaceEnv, getNamespace("base"))
```

As far as I understand it, the reason why there are two "base" environment is that `getNamespace("base")` is used in R as a workaround in order to be able search for objects as explained in Appendix 'Why R searches also the global environment and attached packages' above.  The design is that a package namespace should have all exported **base** objects on the search path _immediately_ following the package's imported objects.  However, because of the historical reasons of having to search also the global environment and the attached packages, we cannot use `baseenv()` for this purpose.  If we did, then the search path would end there, and like breaks a few things as it stands right now.  Because of this, `getNamespace("base")` was introduced as an copy of `baseenv()`, but with the parent environment being the global environment in order to _not_ break the full search path.  If, or rather when, R Core can sort out the last hurdles for having the package search path end immediately after the "import" and the **base** environment, then I imagine there is no longer a need for the special `getNamespace("base")` environment.


## It is not possible to import from the 'base' package

If we would add:

```
importFrom(base, mean)
```

to the `NAMESPACE` file for **teeny**, the package would build, but it
would not _install_.  We would get an installation error:

```
$ R CMD INSTALL teeny_1.0.tar.gz 
* installing to library ‘/path/to/R/x86_64-pc-linux-gnu-library/4.2’
* installing *source* package ‘teeny’ ...
** using staged installation
** R
** byte-compile and prepare package for lazy loading
Error in asNamespace(ns, base.OK = FALSE) : 
  operation not allowed on base namespace
Calls: <Anonymous> ... namespaceImportFrom -> importIntoEnv -> getNamespaceInfo -> asNamespace
ERROR: lazy loading failed for package ‘teeny’
* removing ‘/home/hb/R/x86_64-pc-linux-gnu-library/4.2-CBI-gcc9/teeny’
* restoring previous ‘/home/hb/R/x86_64-pc-linux-gnu-library/4.2-CBI-gcc9/teeny’
```


[roxygen2]: https://cran.r-project.org/package=roxygen2
